---
title: "Chapter 5"
author: "kmm"
date: "February 24, 2016"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(rethinking)
library(ggplot2)
library(GGally)
library(cowplot)
library(ggrepel)
library(dplyr)
library(rstanarm)

set.seed(10)
```

## Waffle House

```{r}
data(WaffleDivorce)
d <- WaffleDivorce

d$MedianAgeMarriage_s <- as.numeric(scale(d$MedianAgeMarriage))
```

## Divorce ~ MedianAgeMarriage 

```{r}
m5.1 <- map2stan(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bA * MedianAgeMarriage_s,
    a ~ dnorm(10, 10),
    bA ~ dnorm(0, 1),
    sigma ~ dcauchy(0, 2)
  ), data = d
)
precis(m5.1)

MAM_seq <- seq(-4, 4, length.out = 30)
mu <- link(m5.1, data = data.frame(MedianAgeMarriage_s = MAM_seq))
mu_PI <- apply(mu, 2, PI)

pred_dat <- data.frame(MAM_seq = MAM_seq,
                       PI_upper = mu_PI[2, ],
                       PI_lower = mu_PI[1, ])

ggplot() +
  geom_abline(intercept = coef(m5.1)["a"],
              slope = coef(m5.1)["bA"]) +
  geom_point(data = d, aes(x = MedianAgeMarriage_s, y = Divorce),
             color = "blue") +
  geom_ribbon(data = pred_dat, aes(x = MAM_seq,
                                   ymax = PI_upper,
                                   ymin = PI_lower),
              fill = "gray50", alpha = 0.2)
```

Try separate for each state

```{r}
d$Loc_ID <- coerce_index(d$Loc)

m5.1_ID <- map2stan(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <-
      a +
      bA * MedianAgeMarriage_s +
      v_Intercept[Loc_ID],
    a ~ dnorm(10, 10),
    bA ~ dnorm(0, 1),
    v_Intercept[Loc_ID] ~ dnorm(0, sigma_Loc_ID),
    sigma_Loc_ID ~ dcauchy(0, 2),
    sigma ~ dcauchy(0, 2)
  ), data = d, iter = 100000,
  control = list(adapt_delta = 0.95)
)

precis(m5.1_ID, digits = 5)
precis(m5.1, digits = 5)

rethinking::compare(m5.1, m5.1_ID)

m5.1_arm <- stan_glm(Divorce ~ MedianAgeMarriage_s, data = d)
m5.1_ID_arm <- stan_lmer(Divorce ~ MedianAgeMarriage_s + (1 | Loc_ID),
                         data = d, adapt_delta = 0.99)
loo1 <- loo(m5.1_arm)
loo2 <- loo(m5.1_ID_arm)
rstanarm::compare(loo1, loo2)

kf1 <- kfold(m5.1_arm, K = 10)
kf2 <- kfold(m5.1_ID_arm, K = 10)
rstanarm::compare(kf2, kf1)

```

## Divorce ~ MedianAgeMarriage + Marriage

```{r}
d$Marriage_s <- as.numeric(scale(d$Marriage))
m5.3 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- intercept + bR * Marriage_s + bA * MedianAgeMarriage_s,
    intercept ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ), data = d
)
precis(m5.3)
plot(precis(m5.3))
```

### Predictor residual plots

Residuals of the predictors regressed against one another. Here residuals for marriage predicted by median age.

```{r}
m5.4 <- map(
  alist(
    Marriage_s ~ dnorm(mu, sigma),
    mu <- intercept + bA * MedianAgeMarriage_s,
    intercept ~ dnorm(0, 10),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ), data = d
)
mu <- coef(m5.4)["intercept"] + coef(m5.4)["bA"] * d$MedianAgeMarriage_s
m_resid <- d$Marriage_s - mu
plot(m_resid, d$Divorce)
```

### Counterfactual plots

Implied predictions of the model. Ask for the predictions for all observed or unobserved data.

```{r}
# Prepare counterfactual data. Change in Marriage_s on predictions
A_avg <- mean(d$MedianAgeMarriage_s)
R_seq <- seq(-3, 3, length.out = 30)
pred_data <- data.frame(Marriage_s = R_seq,
                        MedianAgeMarriage_s = A_avg)
# Compute counterfactual mean divorce (mu)
mu <- link(m5.3, data = pred_data, n = 1e4)
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
# Simulate counterfactual divorce outcomes
R_sim <- sim(m5.3, data = pred_data, n = 1e4)
R_PI <- apply(R_sim, 2, PI)
# Display
plot(Divorce ~ Marriage_s, data = d, type = "n")
mtext("MedianAgeMarriage_s = 0")
lines(R_seq, mu_mean)
shade(mu_PI, R_seq)
shade(R_PI, R_seq)

# Change in MedianAgeMarriage_s on predictions
R_avg <- mean(d$Marriage_s)
A_seq <- seq(-3, 3.5, length.out = 30)
pred_data2 <- data.frame(Marriage_s = R_avg,
                         MedianAgeMarriage_s = A_seq)
# Compute counterfactual mean divorce (mu)
mu <- link(m5.3, data = pred_data2, n = 1e4)
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
# Simulate counterfactual divorce outcomes
A_sim <- sim(m5.3, data = pred_data2, n = 1e4)
A_PI <- apply(A_sim, 2, PI)
# Display
plot(Divorce ~ MedianAgeMarriage_s, data = d, type = "n")
mtext("Marriage_s = 0")
lines(A_seq, mu_mean)
shade(mu_PI, A_seq)
shade(A_PI, A_seq)
```

### Posterior prediction plots

```{r post_pred}
# Call link without new data, so it uses the original data
mu <- link(m5.3)

# Summarize samples across cases
mu_mean <- apply(mu, 2, mean) # Mean value for the sample at each obs
mu_PI <- apply(mu, 2, PI)     # PI for each obs

# Simulate observations
divorce_sim <- sim(m5.3, n = 1e4)
divorce_PI <- apply(divorce_sim, 2, PI)
```

```{r}
# Plot predicted against observed
plot(mu_mean ~ d$Divorce, col = rangi2, ylim = range(mu_PI),
     xlab = "Observed Divorce", ylab = "Predicted Divorce")
abline(a = 0, b = 1, lty = 2)
for (i in 1:nrow(d)) {
  lines(rep(d$Divorce[i], 2), c(mu_PI[1, i], mu_PI[2, i]),
        col = rangi2)
}

d_pred <- data.frame(State = as.character(d$Loc),
                     Divorce = d$Divorce,
                     Predicted = mu_mean,
                     Lower = mu_PI[1, ],
                     Upper = mu_PI[2, ])
ggplot(d_pred, aes(x = Divorce, y = Predicted, label = State)) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point() +
  geom_text_repel(size = 3) +
  geom_linerange(aes(ymin = Lower, ymax = Upper))

d_pred$resid <- d$Divorce - mu_mean
d_pred$Order <- 1:nrow(d_pred)
d_pred <- d_pred %>% arrange(resid)

# Residual plot
divorce_resid <- d$Divorce - mu_mean
o <- order(divorce_resid)
dotchart(divorce_resid[o], labels = d$Loc[o],
         xlim = c(-6, 5), cex = 0.6)
abline(v = 0, col = alpha("black", 0.2))
for (i in 1:nrow(d)) {
  j <- o[i]
  lines(d$Divorce[j] -  c(mu_PI[1, j], mu_PI[2, j]), rep(i, 2))
  points(d$Divorce[j] - c(divorce_PI[1, j], divorce_PI[2, j]),
         rep(i, 2), pch = 3, cex = 0.6, col = "gray")
}
```

## 5.2 Masked relationship

Milk energy vs. Neocortex

```{r}
data(milk)
d <- milk
str(d)
d <- d %>% select(kcal.per.g, neocortex.perc, mass) %>%
  filter(complete.cases(.))
```

```{r map_milk}
m5.5 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ), data = d
)
precis(m5.5, digits = 3)

np_seq <- 0:100
pred_data <- data.frame(neocortex.perc = np_seq)
mu <- link(m5.5, data = pred_data, n = 1e4)
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
```

```{r}
plot(kcal.per.g ~ neocortex.perc, data = d, col = rangi2)
lines(np_seq, mu_mean)
lines(np_seq, mu_PI[1, ], lty = 2)
lines(np_seq, mu_PI[2, ], lty = 2)

d_pred <- data.frame(neocortex.perc = np_seq,
                     mu_mean = mu_mean,
                     Lower = mu_PI[1, ],
                     Upper = mu_PI[2, ])
ggplot() +
  geom_point(data = d, aes(neocortex.perc, kcal.per.g),
             color = "darkblue") +
  geom_line(data = d_pred, aes(neocortex.perc, mu_mean)) +
  geom_ribbon(data = d_pred, aes(x = neocortex.perc,
                                    ymin = Lower,
                                    ymax = Upper),
              fill = "gray50",
              alpha = 0.2) +
  labs(x = "Percent Neocortex", y = "Energy Usage (kcal/g")
```

Milk energy vs. Mass

```{r}
d$logMass <- log(d$mass)
m5.6 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * logMass,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ), data = d
)
precis(m5.6)
```

Milk energy vs. Neocortex + Mass

```{r}
m5.7 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc + bm * logMass,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    bm ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ), data = d
)
precis(m5.7)
```

Counterfactual plots

```{r}
mean_logMass <- mean(log(d$mass))
np_seq <- 0:100
pred_data <- data.frame(
  neocortex.perc = np_seq,
  logMass = mean_logMass
)
mu <- link(m5.7, data = pred_data, n = 1e4)
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)

plot(kcal.per.g ~ neocortex.perc, data = d, type = "n")
lines(np_seq, mu_mean)
lines(np_seq, mu_PI[1, ], lty = 2)
lines(np_seq, mu_PI[2, ], lty = 2)
```

```{r}
mean_neo <- mean(d$neocortex.perc)
np_seq <- seq(-2.5, 4.5, length = 200)
pred_data <- data.frame(
  logMass = np_seq,
  neocortex.perc = mean_neo
)
mu <- link(m5.7, data = pred_data, n = 1e4)
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)

plot(kcal.per.g ~ logMass, data = d, type = "n")
lines(np_seq, mu_mean)
lines(np_seq, mu_PI[1, ], lty = 2)
lines(np_seq, mu_PI[2, ], lty = 2)
```

## Simulate masking

```{r}
set.seed(5)
N <- 100                         # number of cases
rho <- 0.7                       # correlation btw x_pos and x_neg
x_pos <- rnorm(N)                # x_pos as Gaussian
x_neg <- rnorm(N, rho * x_pos,   # x_neg correlated with x_pos
    sqrt(1 - rho ^ 2))
y <- rnorm(N, x_pos - x_neg)     # y equally associated with x_pos, x_neg
d <- data.frame(y, x_pos, x_neg) # bind all together in data frame
pairs(d)

ggscatmat(d)
```

## Simulate multicollinearity

```{r}
set.seed(100)
n <- 100
height <- rnorm(n, 10, 2)
leg_prop <- runif(n, 0.4, 0.5)
left_leg <- leg_prop * height + rnorm(n, 0, 0.02)
right_leg <- leg_prop * height + rnorm(n, 0, 0.02)
d <- data.frame(height, left_leg, right_leg)
```

```{r}
m5.8 <- map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl * left_leg + br * right_leg,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    br ~ dnorm(2, 10),
    sigma ~ dunif(0, 10)
  ), data = d
)
precis(m5.8)
plot(precis(m5.8))
post <- extract.samples(m5.8)
dens(post)
ggplot(post, aes(bl, br)) + geom_point(alpha = 0.2)
dens(post$bl + post$br)
```

```{r leg_stan}
m5.8_s <- map2stan(m5.8)
plot(m5.8_s)
pairs(m5.8_s)
plot(precis(m5.8_s))
precis(m5.8_s)
```

## Milk multicollinearity

```{r}
data(milk)
d <- milk[, c(3, 4, 6)]
names(d) <- c("Energy", "Fat", "Lactose")
d <- d[complete.cases(d), ]
```

```{r}
m5.10 <- map(
  alist(
    Energy ~ dnorm(mu, sigma),
    mu <- a + bF * Fat,
    a ~ dnorm(0.6, 10),
    bF ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ), data = d
)

m5.11 <- map(
  alist(
    Energy ~ dnorm(mu, sigma),
    mu <- a + bL * Lactose,
    a ~ dnorm(0.6, 10),
    bL ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ), data = d
)
coeftab(m5.10, m5.11, digits = 3)
rethinking::compare(m5.10, m5.11)
ggscatmat(d)
```

Start at 5.3.3
